---
title: "Point Win Probability — Replication and SoS Extension"
subtitle: "Harvard Women's Volleyball 2018 (built on Powers, Stancil & Consiglio 2025)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
library(ggplot2)
library(scales)
# Required packages: datavolley, dplyr, tibble, tidyr, lme4, ggrepel, purrr
```

# 1. Setup and Data Loading

```{r config}
randomize_blocker_digger_resp <- FALSE
data_dir <- "2018dvwfiles"   # adjust to your local path
```

```{r load-data}
dvw_files <- list.files(data_dir, pattern = "\\.dvw$", full.names = TRUE)

plays_list <- list()
for (file in dvw_files) {
  data <- try(
    datavolley::dv_read(file, insert_technical_timeouts = FALSE),
    silent = TRUE
  )
  if ("datavolley" %in% class(data)) {
    plays_list[[file]] <- datavolley::plays(data)
  }
}

plays <- dplyr::bind_rows(plays_list) |>
  tibble::as_tibble()

cat("Total rows:", nrow(plays), "\n")
cat("Matches loaded:", length(plays_list), "\n")
```

# 2. Lookup Tables

```{r lookup-tables}
team <- plays |>
  dplyr::filter(!is.na(team_id)) |>
  dplyr::distinct(team_id, team_name = team)

player <- plays |>
  dplyr::filter(!is.na(player_id), player_id != "unknown player") |>
  dplyr::count(player_id, player_name, team_id) |>
  dplyr::arrange(-n) |>
  dplyr::group_by(player_id) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select(player_id, player_name, team_id)

team_lineup_away <- plays |>
  dplyr::select(match_id, set_number, team_id = visiting_team_id,
                dplyr::starts_with("visiting_player_id")) |>
  dplyr::distinct() |>
  tidyr::pivot_longer(cols = dplyr::starts_with("visiting_player_id"),
                      values_to = "player_id") |>
  dplyr::distinct(match_id, set_number, team_id, player_id) |>
  dplyr::filter(player_id != "unkown player")

team_lineup_home <- plays |>
  dplyr::select(match_id, set_number, team_id = home_team_id,
                dplyr::starts_with("home_player_id")) |>
  dplyr::distinct() |>
  tidyr::pivot_longer(cols = dplyr::starts_with("home_player_id"),
                      values_to = "player_id") |>
  dplyr::distinct(match_id, set_number, team_id, player_id) |>
  dplyr::filter(player_id != "unkown player")

team_lineup <- dplyr::bind_rows(team_lineup_away, team_lineup_home)

libero <- plays |>
  dplyr::filter(skill %in% c("Serve", "Reception", "Dig")) |>
  dplyr::count(match_id, set_number, team_id, player_id) |>
  dplyr::anti_join(team_lineup,
                   by = c("match_id", "set_number", "team_id", "player_id")) |>
  dplyr::group_by(match_id, set_number, team_id) |>
  dplyr::arrange(-n) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select(match_id, set_number, team_id, libero_id = player_id)

cat("Teams:", nrow(team), "  Players:", nrow(player), "\n")

# After the §4 cleaning step, with `contact` and `bad_data` already in memory
contact_clean <- contact |>
  dplyr::anti_join(bad_data, by = c("match_id", "point_id"))

# Total cleaned
nrow(contact_clean)
# Should be 20,439

# Breakdown by what gets counted vs not
contact_clean |>
  dplyr::mutate(
    bucket = dplyr::case_when(
      skill %in% c("Serve", "Reception", "Set", "Attack", "Block", "Dig") ~ "six_core_skills",
      point ~ "terminal_point_marker",
      is.na(skill) ~ "no_skill",
      TRUE ~ paste0("other_skill: ", skill)
    )
  ) |>
  dplyr::count(bucket, sort = TRUE)
```

# 3. Markov Chain States

```{r markov-states}
cumpaste <- function(x, .sep = " ") {
  concat <- paste(x, collapse = .sep)
  substring(concat, 1L, cumsum(c(nchar(x[[1L]]), nchar(x[-1L]) + nchar(.sep))))
}

contact <- plays |>
  dplyr::mutate(
    team_id_offense = team_id,
    team_id_defense = ifelse(team_id_offense == home_team_id,
                             visiting_team_id, home_team_id),
    serve_receive = ifelse(team == serving_team, "S", "R"),
    is_volley_end = dplyr::coalesce(
      dplyr::lead(point, 1) | (team_touch_id != dplyr::lead(team_touch_id, 1)),
      TRUE
    ),
    abbrev = dplyr::case_when(
      point ~ "P",
      is.na(skill) ~ NA_character_,
      grepl("Unknown", skill) ~ NA_character_,
      skill == "Serve" ~ "SV",
      skill == "Attack" ~ paste0("A", attack_code),
      TRUE ~ paste0(substring(skill, 1, 1), evaluation_code)
    ),
    player_id_lead_1 = dplyr::lead(player_id, 1),
    skill_lead_1 = dplyr::lead(skill, 1),
    evaluation_code_lead_1 = dplyr::lead(evaluation_code, 1)
  ) |>
  dplyr::filter(!is.na(abbrev), !is.na(serving_team)) |>
  dplyr::group_by(match_id, point_id, team_touch_id, point) |>
  dplyr::mutate(
    state = paste0(serve_receive[1], "_", cumpaste(abbrev, .sep = ".")),
    num_contacts = length(abbrev)
  ) |>
  dplyr::ungroup()

cat("Total contacts:", nrow(contact),
    "  Distinct states:", dplyr::n_distinct(contact$state), "\n")
```

# 4. Data Cleaning

```{r data-cleaning}
bad_data <- contact |>
  dplyr::group_by(state) |>
  dplyr::mutate(n = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::filter(
    (num_contacts > 4) |
    (abbrev %in% c("SV", "A") & !is_volley_end) |
    (n < 10)
  ) |>
  dplyr::distinct(match_id, point_id)

cat("Bad points removed:", nrow(bad_data), "\n")
cat("Contacts remaining:",
    nrow(contact |> dplyr::anti_join(bad_data, by = c("match_id", "point_id"))),
    "\n")
```

# 5. Transition Matrix and Sideout Probability

```{r transitions}
state_transition <- contact |>
  dplyr::anti_join(bad_data, by = c("match_id", "point_id")) |>
  dplyr::count(state, next_state = dplyr::lead(state, 1)) |>
  dplyr::mutate(
    next_state = dplyr::case_when(
      state == "R_P" ~ "R_P",
      state == "S_P" ~ "S_P",
      TRUE ~ next_state
    )
  )

state_transition_wide <- state_transition |>
  dplyr::bind_rows(tibble::tibble(state = "S_SV", next_state = "S_SV", n = 0)) |>
  dplyr::group_by(state, next_state) |>
  dplyr::summarize(n = sum(n), .groups = "drop") |>
  dplyr::group_by(state) |>
  dplyr::transmute(state, next_state, prob = n / sum(n)) |>
  dplyr::arrange(next_state) |>
  tidyr::pivot_wider(names_from = next_state, values_from = prob, values_fill = 0) |>
  dplyr::arrange(state)

state_transition_matrix <- state_transition_wide |>
  tibble::column_to_rownames("state") |>
  as.matrix()
```

```{r sideout-prob}
state_transition_matrix_limit <- state_transition_matrix
for (i in 1:100) {
  state_transition_matrix_limit <- state_transition_matrix_limit %*% state_transition_matrix
}

sideout_prob <- state_transition_matrix_limit[, c("S_P", "R_P")] |>
  tibble::as_tibble() |>
  tibble::add_column(state = rownames(state_transition_matrix), .before = 1) |>
  dplyr::transmute(state, sideout_prob = R_P)

win_prob_from_dig <- sideout_prob |>
  dplyr::filter(nchar(state) == 4, substring(state, 1, 3) == "R_D") |>
  dplyr::transmute(eval = substring(state, 4), win_prob = sideout_prob)

contact_sideout_prob <- contact |>
  dplyr::left_join(sideout_prob, by = "state") |>
  dplyr::mutate(sideout_prob_lead_1 = dplyr::lead(sideout_prob, 1))

cat("Sideout from serve (S_SV):",
    round(sideout_prob$sideout_prob[sideout_prob$state == "S_SV"], 4), "\n")
```

# 5b. Pass Quality Ladder

The cleanest face-validity check for the Markov chain output: sideout probability
should rise monotonically with reception evaluation, and each step up the
reception scale should be worth a meaningful chunk of win probability.

```{r pass-ladder, fig.height = 4}
state_freq <- contact |>
  dplyr::anti_join(bad_data, by = c("match_id", "point_id")) |>
  dplyr::count(state, name = "n")

baseline_serve <- sideout_prob$sideout_prob[sideout_prob$state == "S_SV"]

pass_ladder <- sideout_prob |>
  dplyr::filter(state %in% c("S_SV", "R_R#", "R_R+", "R_R!", "R_R-", "R_R=")) |>
  dplyr::left_join(state_freq, by = "state") |>
  dplyr::mutate(
    description = dplyr::case_when(
      state == "S_SV" ~ "Baseline (just-served)",
      state == "R_R#" ~ "Perfect pass",
      state == "R_R+" ~ "Good pass",
      state == "R_R!" ~ "Medium pass",
      state == "R_R-" ~ "Poor pass",
      state == "R_R=" ~ "Reception error / ace"
    )
  ) |>
  dplyr::arrange(dplyr::desc(sideout_prob))

pass_ladder |>
  dplyr::filter(state != "S_SV") |>
  dplyr::mutate(state = factor(state,
                               levels = c("R_R=", "R_R-", "R_R!", "R_R+", "R_R#"))) |>
  ggplot(aes(x = state, y = sideout_prob)) +
  geom_col(fill = "#3b5bdb", width = 0.6) +
  geom_hline(yintercept = baseline_serve, linetype = "dashed", color = "grey40") +
  geom_text(aes(label = sprintf("%.3f", sideout_prob)),
            vjust = -0.5, size = 3.5) +
  annotate("text", x = 1, y = baseline_serve + 0.02, hjust = 0,
           label = sprintf("Baseline (just-served) = %.3f", baseline_serve),
           color = "grey30", size = 3.2) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, 0.75), expand = c(0, 0)) +
  scale_x_discrete(labels = c("R= (ace)", "R- (poor)", "R! (medium)",
                              "R+ (good)", "R# (perfect)")) +
  labs(
    title = "Pass quality drives sideout probability",
    subtitle = sprintf("Each step up the reception scale is worth %d-%d percentage points of win probability",
                       5, 13),
    x = NULL, y = "Receiver point-win probability"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        plot.title.position = "plot")
```

The ladder is monotone and the gaps are large: a perfect pass is worth roughly
13 percentage points of win probability over a poor pass. This is the simplest
visualization of why per-contact PG is informative — the per-contact win
probability changes are large enough to support meaningful player evaluation.

# 6. Serve and Reception PG (raw)

```{r serve-data}
serve_data <- contact_sideout_prob |>
  dplyr::filter(
    skill == "Serve",
    skill_lead_1 == "Reception" | evaluation_code == "=",
    !is.na(sideout_prob_lead_1),
    player_id != "unknown player",
    !is.na(home_team_id),
    !is.na(visiting_team_id)
  ) |>
  dplyr::mutate(
    server_id = player_id,
    receiver_id = player_id_lead_1,
    is_error = evaluation_code == "=",
    no_error = !is_error,
    exp_no_error = weighted.mean(sideout_prob_lead_1, w = !is_error),
    pg_is_error = ifelse(is_error, 1, exp_no_error) - sideout_prob,
    pg_no_error = ifelse(is_error, 0, sideout_prob_lead_1 - exp_no_error)
  )
```

```{r pg-serve-raw}
pg_serve <- serve_data |>
  dplyr::select(-player_name, -team_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name, team_id),
                   by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    pg = -mean(pg_is_error + pg_no_error),
    n = dplyr::n(),
    pct_is_error = mean(is_error),
    pg_is_error = -mean(pg_is_error),
    n_no_error = sum(no_error),
    pct_ace = weighted.mean(evaluation_code == "#", w = no_error),
    pg_no_error = -mean(pg_no_error),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg)
```

```{r pg-reception-raw}
pg_reception <- serve_data |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = receiver_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name, team_id),
                   by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n_reception = sum(no_error),
    pct_ace = weighted.mean(evaluation_code == "#", w = no_error),
    pg_reception = weighted.mean(pg_no_error, w = no_error),
    .groups = "drop"
  )
```

# 7. Attack Outcome Tree

```{r attack-prep}
attack <- contact_sideout_prob |>
  dplyr::left_join(libero,
                   by = c("match_id", "set_number", "home_team_id" = "team_id")) |>
  dplyr::rename(home_libero_id = libero_id) |>
  dplyr::left_join(libero,
                   by = c("match_id", "set_number", "visiting_team_id" = "team_id")) |>
  dplyr::rename(visiting_libero_id = libero_id) |>
  dplyr::mutate(
    is_error = evaluation_code == "=",
    is_no_error = !is_error,
    is_block = !is_error & dplyr::coalesce(dplyr::lead(skill, 1) == "Block", FALSE),
    is_no_block = !is_error & !is_block,
    is_block_error = is_block & dplyr::lead(evaluation_code, 1) == "=",
    is_block_no_error = is_block & !is_block_error,
    is_block_return = is_block_no_error & (
      dplyr::lead(evaluation_code, 1) == "#" |
      dplyr::coalesce(dplyr::lead(skill, 2) == "Dig", FALSE) &
        team_id == dplyr::lead(team_id, 2)
    ),
    is_block_through = is_block_no_error & (
      dplyr::coalesce(dplyr::lead(skill, 2) == "Dig", FALSE) &
        team_id != dplyr::lead(team_id, 2)
    ),
    outcome_eval = dplyr::case_when(
      is_error ~ "=",
      !is_error & !is_block & dplyr::lead(skill, 1) %in% c("Dig", NA) ~
        dplyr::coalesce(dplyr::lead(evaluation_code, 1), "="),
      is_block_error ~ "=",
      is_block & !is_block_error & !is_block_return &
        dplyr::lead(skill, 2) %in% c("Dig", NA) ~
        dplyr::coalesce(dplyr::lead(evaluation_code, 2), "="),
      is_block_return & dplyr::lead(skill, 2) %in% c("Dig", NA) ~
        dplyr::coalesce(dplyr::lead(evaluation_code, 2), "=")
    ),
    outcome_team = ifelse(is_error | is_block_return, "offense", "defense"),
    defense_setter_position = ifelse(team_id == home_team_id,
                                     visiting_setter_position,
                                     home_setter_position),
    defense_player_id1 = ifelse(team_id == home_team_id, visiting_player_id1, home_player_id1),
    defense_player_id2 = ifelse(team_id == home_team_id, visiting_player_id2, home_player_id2),
    defense_player_id3 = ifelse(team_id == home_team_id, visiting_player_id3, home_player_id3),
    defense_player_id4 = ifelse(team_id == home_team_id, visiting_player_id4, home_player_id4),
    defense_player_id5 = ifelse(team_id == home_team_id, visiting_player_id5, home_player_id5),
    defense_player_id6 = ifelse(team_id == home_team_id, visiting_player_id6, home_player_id6),
    defense_libero_id = ifelse(team_id == home_team_id, visiting_libero_id, home_libero_id),
    offense_setter_position = ifelse(team_id == visiting_team_id,
                                     visiting_setter_position,
                                     home_setter_position),
    offense_player_id1 = ifelse(team_id == visiting_team_id, visiting_player_id1, home_player_id1),
    offense_player_id2 = ifelse(team_id == visiting_team_id, visiting_player_id2, home_player_id2),
    offense_player_id3 = ifelse(team_id == visiting_team_id, visiting_player_id3, home_player_id3),
    offense_player_id4 = ifelse(team_id == visiting_team_id, visiting_player_id4, home_player_id4),
    offense_player_id5 = ifelse(team_id == visiting_team_id, visiting_player_id5, home_player_id5),
    offense_player_id6 = ifelse(team_id == visiting_team_id, visiting_player_id6, home_player_id6),
    offense_libero_id = ifelse(team_id == visiting_team_id, visiting_libero_id, home_libero_id),

    defense_back_right_player_id = dplyr::case_when(
      defense_setter_position == 1 ~ defense_player_id1,
      defense_setter_position == 2 ~ defense_player_id5,
      defense_setter_position == 3 ~ defense_player_id6,
      defense_setter_position == 4 ~ defense_player_id1,
      defense_setter_position == 5 ~ defense_player_id5,
      defense_setter_position == 6 ~ defense_player_id6
    ),
    defense_front_right_player_id = dplyr::case_when(
      defense_setter_position == 1 ~ defense_player_id4,
      defense_setter_position == 2 ~ defense_player_id2,
      defense_setter_position == 3 ~ defense_player_id3,
      defense_setter_position == 4 ~ defense_player_id4,
      defense_setter_position == 5 ~ defense_player_id2,
      defense_setter_position == 6 ~ defense_player_id3
    ),
    defense_front_middle_player_id = dplyr::case_when(
      defense_setter_position == 1 ~ defense_player_id3,
      defense_setter_position == 2 ~ defense_player_id4,
      defense_setter_position == 3 ~ defense_player_id2,
      defense_setter_position == 4 ~ defense_player_id3,
      defense_setter_position == 5 ~ defense_player_id4,
      defense_setter_position == 6 ~ defense_player_id2
    ),
    defense_front_left_player_id = dplyr::case_when(
      defense_setter_position == 1 ~ defense_player_id2,
      defense_setter_position == 2 ~ defense_player_id3,
      defense_setter_position == 3 ~ defense_player_id4,
      defense_setter_position == 4 ~ defense_player_id2,
      defense_setter_position == 5 ~ defense_player_id3,
      defense_setter_position == 6 ~ defense_player_id4
    ),
    defense_back_left_player_id = defense_libero_id,
    defense_back_middle_player_id = dplyr::case_when(
      defense_setter_position == 1 ~ defense_player_id5,
      defense_setter_position == 2 ~ defense_player_id6,
      defense_setter_position == 3 ~ defense_player_id1,
      defense_setter_position == 4 ~ defense_player_id5,
      defense_setter_position == 5 ~ defense_player_id6,
      defense_setter_position == 6 ~ defense_player_id1
    ),

    offense_back_right_player_id = dplyr::case_when(
      offense_setter_position == 1 ~ offense_player_id1,
      offense_setter_position == 2 ~ offense_player_id5,
      offense_setter_position == 3 ~ offense_player_id6,
      offense_setter_position == 4 ~ offense_player_id1,
      offense_setter_position == 5 ~ offense_player_id5,
      offense_setter_position == 6 ~ offense_player_id6
    ),
    offense_front_right_player_id = dplyr::case_when(
      offense_setter_position == 1 ~ offense_player_id4,
      offense_setter_position == 2 ~ offense_player_id2,
      offense_setter_position == 3 ~ offense_player_id3,
      offense_setter_position == 4 ~ offense_player_id4,
      offense_setter_position == 5 ~ offense_player_id2,
      offense_setter_position == 6 ~ offense_player_id3
    ),
    offense_front_middle_player_id = dplyr::case_when(
      offense_setter_position == 1 ~ offense_player_id3,
      offense_setter_position == 2 ~ offense_player_id4,
      offense_setter_position == 3 ~ offense_player_id2,
      offense_setter_position == 4 ~ offense_player_id3,
      offense_setter_position == 5 ~ offense_player_id4,
      offense_setter_position == 6 ~ offense_player_id2
    ),
    offense_front_left_player_id = dplyr::case_when(
      offense_setter_position == 1 ~ offense_player_id2,
      offense_setter_position == 2 ~ offense_player_id3,
      offense_setter_position == 3 ~ offense_player_id4,
      offense_setter_position == 4 ~ offense_player_id2,
      offense_setter_position == 5 ~ offense_player_id3,
      offense_setter_position == 6 ~ offense_player_id4
    ),
    offense_back_left_player_id = offense_libero_id,
    offense_back_middle_player_id = dplyr::case_when(
      offense_setter_position == 1 ~ offense_player_id5,
      offense_setter_position == 2 ~ offense_player_id6,
      offense_setter_position == 3 ~ offense_player_id1,
      offense_setter_position == 4 ~ offense_player_id5,
      offense_setter_position == 5 ~ offense_player_id6,
      offense_setter_position == 6 ~ offense_player_id1
    ),

    player_position = dplyr::case_when(
      player_id == offense_front_left_player_id ~ "OH",
      player_id == offense_front_middle_player_id ~ "MB",
      player_id == offense_front_right_player_id ~ "RS",
      player_id %in% c(offense_back_left_player_id,
                       offense_back_middle_player_id,
                       offense_back_right_player_id) ~ "BR"
    ),
    player_zone = dplyr::case_when(
      player_id == offense_back_left_player_id ~ "BL",
      player_id == offense_back_middle_player_id ~ "BM",
      player_id == offense_back_right_player_id ~ "BR",
      player_id == offense_front_left_player_id ~ "FL",
      player_id == offense_front_middle_player_id ~ "FM",
      player_id == offense_front_right_player_id ~ "FR",
      TRUE ~ "BL"
    ),
    setter_id = ifelse(dplyr::lag(skill, 1) == "Set", dplyr::lag(player_id, 1), NA),
    blocker_id = ifelse(dplyr::lead(skill, 1) == "Block", dplyr::lead(player_id, 1), NA),
    digger_id = dplyr::case_when(
      dplyr::lead(skill, 1) == "Dig" ~ dplyr::lead(player_id, 1),
      dplyr::lead(skill, 1) == "Block" & dplyr::lead(skill, 2) == "Dig" ~
        dplyr::lead(player_id, 2),
      TRUE ~ NA
    )
  )
```

```{r volley-and-resp}
volley_start <- attack |>
  dplyr::group_by(match_id, point_id, team_touch_id) |>
  dplyr::filter(skill != "Block") |>
  dplyr::slice(1) |>
  dplyr::transmute(match_id, point_id, team_touch_id,
                   volley_start = paste0(substring(skill, 1, 1), evaluation_code))

blocker_responsibility <- attack |>
  dplyr::mutate(
    blocker_resp = ifelse(dplyr::lead(skill, 1) == "Block",
                          dplyr::lead(player_zone, 1), NA)
  ) |>
  dplyr::count(attack_code, blocker_resp) |>
  dplyr::filter(!is.na(blocker_resp)) |>
  dplyr::group_by(attack_code) |>
  dplyr::arrange(-n) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select(attack_code, blocker_resp)

digger_responsibility <- attack |>
  dplyr::mutate(
    digger_resp = ifelse(dplyr::lead(skill, 1) == "Dig",
                         dplyr::lead(player_zone, 1), NA)
  ) |>
  dplyr::count(attack_code, end_zone, digger_resp) |>
  dplyr::filter(!is.na(digger_resp), !is.na(attack_code)) |>
  dplyr::group_by(attack_code, end_zone) |>
  dplyr::arrange(-n) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select(attack_code, end_zone, digger_resp)
```

```{r attack-data}
attack_data <- attack |>
  dplyr::mutate(
    skill_lead_1 = dplyr::lead(skill, 1),
    skill_lead_2 = dplyr::lead(skill, 2)
  ) |>
  dplyr::filter(
    skill_lead_1 %in% c("Block", "Dig", NA),
    skill_lead_1 %in% c("Dig", NA) | skill_lead_2 %in% c("Dig", NA),
    skill == "Attack",
    !is_block_no_error | xor(is_block_return, is_block_through)
  ) |>
  dplyr::left_join(volley_start, by = c("match_id", "point_id", "team_touch_id")) |>
  dplyr::left_join(blocker_responsibility, by = "attack_code") |>
  dplyr::left_join(digger_responsibility, by = c("attack_code", "end_zone")) |>
  dplyr::left_join(win_prob_from_dig, by = c("outcome_eval" = "eval")) |>
  dplyr::filter(substring(volley_start, 1, 1) %in% c("D", "R", "F")) |>
  dplyr::mutate(
    blocker_resp = ifelse(
      test = rep(randomize_blocker_digger_resp, length = dplyr::n()),
      yes = sample(c("FL", "FM", "FR"), size = dplyr::n(), replace = TRUE),
      no = blocker_resp
    ),
    digger_resp = ifelse(
      test = rep(randomize_blocker_digger_resp, length = dplyr::n()),
      yes = sample(c("BL", "BM", "BR"), size = dplyr::n(), replace = TRUE),
      no = blocker_resp
    ),
    blocker_id = dplyr::case_when(
      !is.na(blocker_id) ~ blocker_id,
      blocker_resp == "FL" ~ defense_front_left_player_id,
      blocker_resp == "FM" ~ defense_front_middle_player_id,
      blocker_resp == "FR" ~ defense_front_right_player_id
    ),
    digger_id = dplyr::case_when(
      !is.na(digger_id) ~ digger_id,
      digger_resp == "FL" ~ defense_front_left_player_id,
      digger_resp == "FM" ~ defense_front_middle_player_id,
      digger_resp == "FR" ~ defense_front_right_player_id,
      digger_resp == "BL" ~ defense_back_left_player_id,
      digger_resp == "BM" ~ defense_back_middle_player_id,
      digger_resp == "BR" ~ defense_back_right_player_id
    ),
    win_prob = ifelse(outcome_team == "offense", win_prob, 1 - win_prob)
  )
```

```{r expected-win-prob}
exp_win_prob <- attack_data |>
  dplyr::filter(!is.na(win_prob), !is.na(attack_code)) |>
  dplyr::group_by(volley_start, attack_code) |>
  dplyr::summarize(
    n = dplyr::n(),
    exp = mean(win_prob),
    exp_error = 0,
    exp_no_error = weighted.mean(win_prob, w = is_no_error),
    exp_block = weighted.mean(win_prob, w = is_block),
    exp_no_block = weighted.mean(win_prob, w = is_no_block),
    exp_block_error = 1,
    exp_block_no_error = weighted.mean(win_prob, w = is_block_no_error),
    exp_block_return = weighted.mean(win_prob, w = is_block_return),
    exp_block_through = weighted.mean(win_prob, w = is_block_through),
    .groups = "drop"
  ) |>
  dplyr::filter(n >= 10)

attack_data_with_pg <- attack_data |>
  dplyr::filter(
    player_id != "unknown player",
    !is.na(home_team_id), !is.na(visiting_team_id), !is.na(win_prob)
  ) |>
  dplyr::inner_join(exp_win_prob, by = c("volley_start", "attack_code")) |>
  dplyr::mutate(
    pg_is_error = dplyr::case_when(
      is_error ~ exp_error,
      is_no_error ~ exp_no_error
    ) - exp,
    pg_is_block = dplyr::case_when(
      is_block ~ exp_block,
      is_no_block ~ exp_no_block,
      TRUE ~ exp_no_error
    ) - exp_no_error,
    pg_is_block_error = dplyr::case_when(
      is_block_error ~ exp_block_error,
      is_block_no_error ~ exp_block_no_error,
      TRUE ~ exp_block
    ) - exp_block,
    pg_is_block_return = dplyr::case_when(
      is_block_return ~ exp_block_return,
      is_block_through ~ exp_block_through,
      TRUE ~ exp_block_no_error
    ) - exp_block_no_error,
    pg_no_block = ifelse(is_no_block, win_prob - exp_no_block, 0),
    pg_block_return = ifelse(is_block_return, win_prob - exp_block_return, 0),
    pg_block_through = ifelse(is_block_through, win_prob - exp_block_through, 0)
  )

cat("Attack rows with PG:", nrow(attack_data_with_pg), "\n")
```

# 8. Per-Skill Raw PG Aggregations

```{r pg-attack-raw}
pg_attack <- attack_data_with_pg |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name, player_position) |>
  dplyr::summarize(
    n = dplyr::n(),
    pct_is_error = mean(is_error),
    pg_is_error = mean(pg_is_error),
    n_no_error = sum(is_no_error),
    pct_is_block = weighted.mean(is_block, w = is_no_error),
    pg_is_block = mean(pg_is_block),
    n_block = sum(is_block),
    pct_is_block_error = weighted.mean(is_block_error, w = is_block),
    pg_is_block_error = mean(pg_is_block_error),
    n_block_no_error = sum(is_block_no_error),
    pct_is_block_through = weighted.mean(is_block_through, w = is_block_no_error),
    pg_is_block_return = mean(pg_is_block_return),
    n_no_block = sum(is_no_block),
    pct_kill_no_block = weighted.mean(win_prob > 0.99, w = is_no_block),
    pg_no_block = mean(pg_no_block),
    n_block_through = sum(is_block_through),
    pct_kill_block_through = weighted.mean(win_prob > 0.99, w = is_block_through),
    pg_block_through = mean(pg_block_through),
    n_block_return = sum(is_block_return),
    pct_stuff = weighted.mean(win_prob < 0.01, w = is_block_return),
    pg_block_return = mean(pg_block_return),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pg = pg_is_error + pg_is_block + pg_is_block_error + pg_is_block_return +
         pg_no_block + pg_block_return + pg_block_through
  ) |>
  dplyr::select(player_id, player_name, team_name, player_position, pg,
                dplyr::everything()) |>
  dplyr::arrange(-pg)
```

```{r pg-set-raw}
pg_set <- attack_data_with_pg |>
  dplyr::select(-player_id, -player_name) |>
  dplyr::rename(player_id = setter_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name), by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = dplyr::n(),
    pct_is_error = mean(is_error),
    pg_is_error = mean(pg_is_error),
    n_no_error = sum(is_no_error),
    pg_is_block = mean(pg_is_block),
    n_block = sum(is_block),
    pg_is_block_error = mean(pg_is_block_error),
    n_block_no_error = sum(is_block_no_error),
    pg_is_block_return = mean(pg_is_block_return),
    n_no_block = sum(is_no_block),
    pg_no_block = mean(pg_no_block),
    n_block_through = sum(is_block_through),
    pg_block_through = mean(pg_block_through),
    n_block_return = sum(is_block_return),
    pg_block_return = mean(pg_block_return),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pg = pg_is_error + pg_is_block + pg_is_block_error + pg_is_block_return +
         pg_no_block + pg_block_return + pg_block_through
  ) |>
  dplyr::select(player_id, player_name, team_name, pg, dplyr::everything()) |>
  dplyr::arrange(-pg)
```

```{r pg-block-raw}
pg_block <- attack_data_with_pg |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = blocker_id) |>
  dplyr::left_join(player, by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = sum(is_no_error),
    pct_is_block = weighted.mean(is_block, w = is_no_error),
    pg_is_block = -weighted.mean(pg_is_block, w = is_no_error),
    n_block = sum(is_block),
    pg_is_block_error = -weighted.mean(pg_is_block_error, w = is_no_error),
    n_block_no_error = sum(is_block_no_error),
    pg_is_block_return = -weighted.mean(pg_is_block_return, w = is_no_error),
    n_no_block = sum(is_no_block),
    pg_no_block = -weighted.mean(pg_no_block, w = is_no_error) / 2,
    n_block_through = sum(is_block_through),
    pg_block_through = -weighted.mean(pg_block_through, w = is_no_error) / 2,
    n_block_return = sum(is_block_return),
    pg_block_return = -weighted.mean(pg_block_return, w = is_no_error),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pg = pg_is_block + pg_is_block_error + pg_is_block_return +
         pg_no_block + pg_block_return + pg_block_through
  ) |>
  dplyr::select(player_id, player_name, team_name, pg, dplyr::everything()) |>
  dplyr::arrange(-pg)
```

```{r pg-dig-raw}
pg_dig <- attack_data_with_pg |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = digger_id) |>
  dplyr::left_join(player, by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n_dig = sum(is_no_block + is_block_through),
    pct_kill = weighted.mean(win_prob > 0.99, w = is_no_block + is_block_through),
    pg_dig = -weighted.mean(
      x = pg_no_block + pg_block_through,
      w = is_no_block + is_block_through
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_dig)
```

```{r pg-overall-raw}
pg_pass <- pg_reception |>
  dplyr::left_join(pg_dig, by = c("player_id", "player_name", "team_name")) |>
  dplyr::mutate(
    n_reception = dplyr::coalesce(n_reception, 0),
    n_dig = dplyr::coalesce(n_dig, 0),
    n = n_reception + n_dig,
    pg = (n_reception * pg_reception + n_dig * pg_dig) / n
  ) |>
  dplyr::select(player_id, player_name, team_name, n, pg, dplyr::everything()) |>
  dplyr::arrange(-pg)

sets_played <- contact_sideout_prob |>
  dplyr::filter(!is.na(player_id)) |>
  dplyr::count(player_id, match_id, set_number) |>
  dplyr::count(player_id) |>
  dplyr::rename(sets_played = n)

id_columns <- c("player_id", "player_name", "team_name")

pg_overall <- sets_played |>
  dplyr::full_join(
    pg_serve |> dplyr::mutate(pg = n * pg) |>
      dplyr::select(dplyr::all_of(id_columns), n_serve = n, pg_serve = pg),
    by = "player_id"
  ) |>
  dplyr::full_join(
    pg_pass |> dplyr::mutate(pg = n * pg) |>
      dplyr::select(dplyr::all_of(id_columns), n_pass = n, pg_pass = pg),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_set |> dplyr::mutate(pg = n * pg) |>
      dplyr::select(dplyr::all_of(id_columns), n_set = n, pg_set = pg),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_attack |> dplyr::mutate(pg = n * pg) |>
      dplyr::group_by_at(id_columns) |>
      dplyr::summarize(n_attack = sum(n), pg_attack = sum(pg), .groups = "drop"),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_block |> dplyr::mutate(pg = n * pg) |>
      dplyr::select(dplyr::all_of(id_columns), n_block = n, pg_block = pg),
    by = id_columns
  ) |>
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("^sets_played|^n_|^pg_")),
    function(x) tidyr::replace_na(x, 0)
  )

pg_overall_per_set <- pg_overall |>
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("^n_|^pg_")),
    function(x, denominator) x / denominator,
    denominator = pg_overall$sets_played
  ) |>
  dplyr::filter(sets_played > 0) |>
  dplyr::mutate(
    n = n_serve + n_pass + n_set + n_attack + n_block,
    pg = pg_serve + pg_pass + pg_set + pg_attack + pg_block
  ) |>
  dplyr::select(dplyr::all_of(id_columns), sets_played, n, pg,
                dplyr::everything()) |>
  dplyr::arrange(-pg)
```

# 8b. Per-Skill PG Distributions

Distribution of player season-average Points Gained per contact, by skill. Each
observation is one player's season-long average PG when performing this skill;
the figure replicates Powers et al. Figure 7 on Harvard 2018 data. Setters cluster
tightest near zero (per-contact win-probability change is small for every set);
attackers spread widest (terminal outcomes generate the largest swings). The
distributions are roughly mean-zero by construction.

```{r table1-counts}
table1_counts <- contact |>
  dplyr::anti_join(bad_data, by = c("match_id", "point_id")) |>
  dplyr::filter(skill %in% c("Serve", "Reception", "Set",
                             "Attack", "Block", "Dig")) |>
  dplyr::count(skill) |>
  dplyr::mutate(skill = factor(skill, levels = c("Serve", "Reception", "Set",
                                                  "Attack", "Block", "Dig"))) |>
  dplyr::arrange(skill)

knitr::kable(table1_counts,
             caption = "Skill contacts in cleaned dataset (Table 1 source)",
             col.names = c("Skill", "Contacts"))

cat("Total:", sum(table1_counts$n), "\n")
```

```{r pg-distribution-data}
pg_per_contact_long <- dplyr::bind_rows(
  pg_serve |>
    dplyr::filter(n >= 20) |>
    dplyr::transmute(player_name, team_name, pg = pg, skill = "Serve"),
  pg_set |>
    dplyr::filter(n >= 50) |>
    dplyr::transmute(player_name, team_name, pg = pg, skill = "Set"),
  pg_attack |>
    dplyr::filter(n >= 30) |>
    dplyr::transmute(player_name, team_name, pg = pg, skill = "Attack"),
  pg_reception |>
    dplyr::filter(n_reception >= 30) |>
    dplyr::transmute(player_name, team_name, pg = pg_reception,
                     skill = "Reception"),
  pg_dig |>
    dplyr::filter(n_dig >= 30) |>
    dplyr::transmute(player_name, team_name, pg = pg_dig, skill = "Dig"),
  pg_block |>
    dplyr::filter(n >= 30) |>
    dplyr::transmute(player_name, team_name, pg = pg, skill = "Block")
) |>
  dplyr::mutate(
    skill = factor(skill, levels = c("Serve", "Set", "Attack",
                                     "Reception", "Dig", "Block"))
  )
```

```{r pg-distribution-plot, fig.height = 5}
ggplot(pg_per_contact_long, aes(x = pg)) +
  geom_histogram(bins = 20, fill = "#3b5bdb", color = "white",
                 linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.4) +
  facet_wrap(~ skill, scales = "free_y", nrow = 2) +
  coord_cartesian(xlim = c(-0.10, 0.10)) +
  scale_x_continuous(labels = percent_format(accuracy = 1),
                     breaks = seq(-0.10, 0.10, 0.05)) +
  scale_y_continuous(breaks = function(x) seq(0, ceiling(max(x)), by = 1)) +
  labs(
    title = "Distribution of player Points Gained per contact, by skill",
    subtitle = "Each observation is one player's season-average PG per contact (replicates Powers et al. Figure 7)",
    x = "Player average Points Gained per contact",
    y = "Number of players"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank(),
        plot.title.position = "plot")
```

```{r pg-range-table}
pg_per_contact_summary <- pg_per_contact_long |>
  dplyr::group_by(skill) |>
  dplyr::summarize(
    n_players = dplyr::n(),
    mean_pg = mean(pg, na.rm = TRUE),
    sd_pg = sd(pg, na.rm = TRUE),
    p05 = quantile(pg, 0.05, na.rm = TRUE),
    p95 = quantile(pg, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(sd_pg)

pg_per_contact_summary |>
  knitr::kable(
    caption = "Player-level PG per contact: spread by skill (replicates Powers et al. Figure 7)",
    digits = 4
  )
```


# 9. Strength of Schedule Adjustment — Serve and Reception

We adapt Powers et al. §3.2.1 to single-conference data by dropping the
conference layer:

$$Y_i \sim \mathrm{Normal}\!\left(\alpha^{*} + \tau_{t} + \pi_{p}
                                   + \tilde{\tau}_{\tilde t} + \rho_{r},\;
                                   \sigma^2_\epsilon\right)$$

The conference effect is absorbed into the intercept $\alpha^{*}$. Per-contact
opponent quality is computed via `predict()` with only opponent-side random
effects active.

```{r sos-serve-prep}
sos_threshold_sv <- 20

serve_data_sos <- serve_data |>
  dplyr::group_by(server_id) |>
  dplyr::mutate(count_sv = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(server_id_model = ifelse(count_sv < sos_threshold_sv,
                                         "0", server_id)) |>
  dplyr::group_by(receiver_id) |>
  dplyr::mutate(count_rc = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(receiver_id_model = ifelse(count_rc < sos_threshold_sv,
                                           "0", receiver_id))

cat("Non-error pairs for serve LMM:", sum(serve_data_sos$no_error), "\n")
cat("Distinct teams (off/def):",
    dplyr::n_distinct(serve_data_sos$team_id_offense), "/",
    dplyr::n_distinct(serve_data_sos$team_id_defense), "\n")
```

```{r sos-serve-fit}
serve_model <- serve_data_sos |>
  dplyr::filter(no_error) |>
  with(
    lme4::lmer(
      pg_no_error ~ (1 | team_id_offense) + (1 | team_id_defense) +
                    (1 | server_id_model) + (1 | receiver_id_model),
      control = lme4::lmerControl(calc.derivs = FALSE, optimizer = "bobyqa")
    )
  )

serve_var_components <- lme4::VarCorr(serve_model) |>
  tibble::as_tibble() |>
  dplyr::select(grp, sd = sdcor)

serve_var_components |>
  knitr::kable(caption = "Serve/reception SoS — variance components", digits = 4)
```

```{r sos-serve-predict}
# Following Powers et al. — opponent-side prediction only
serve_data_adj <- serve_data_sos |>
  dplyr::mutate(
    sos_no_error_server = predict(
      serve_model, newdata = serve_data_sos,
      re.form = ~ (1 | team_id_defense) + (1 | receiver_id_model),
      allow.new.levels = TRUE
    ),
    sos_no_error_receiver = predict(
      serve_model, newdata = serve_data_sos,
      re.form = ~ (1 | team_id_offense) + (1 | server_id_model),
      allow.new.levels = TRUE
    )
  )
```

```{r sos-pg-serve}
pg_serve_adj <- serve_data_adj |>
  dplyr::select(-player_name, -team_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name, team_id),
                   by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = dplyr::n(),
    pg_raw = -mean(pg_is_error + pg_no_error),
    pg_adj = -mean(pg_is_error + pg_no_error - sos_no_error_server),
    sos_avg = mean(sos_no_error_server),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_adj)

pg_reception_adj <- serve_data_adj |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = receiver_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name, team_id),
                   by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n_reception = sum(no_error),
    pg_reception_raw = weighted.mean(pg_no_error, w = no_error),
    pg_reception_adj = weighted.mean(pg_no_error - sos_no_error_receiver,
                                     w = no_error),
    sos_avg = weighted.mean(sos_no_error_receiver, w = no_error),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_reception_adj)

pg_serve_adj |>
  dplyr::filter(n >= 20) |>
  dplyr::filter(dplyr::row_number() <= 10 | team_name == "Harvard University") |>
  knitr::kable(caption = "Serve PG: top 10 servers + all Harvard servers (full table in appendix)",
               digits = 4)
```

```{r sos-serve-scatter, fig.height = 4}
sos_sv_rec_data <- dplyr::bind_rows(
  pg_serve_adj |>
    dplyr::filter(n >= 20) |>
    dplyr::transmute(player_name, team_name, pg_raw, pg_adj, role = "Serve"),
  pg_reception_adj |>
    dplyr::filter(n_reception >= 30) |>
    dplyr::transmute(player_name, team_name,
                     pg_raw = pg_reception_raw,
                     pg_adj = pg_reception_adj,
                     role = "Reception")
) |>
  dplyr::mutate(is_harvard = team_name == "Harvard University")

ggplot(sos_sv_rec_data, aes(x = pg_raw, y = pg_adj)) +
  geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey85", linewidth = 0.3) +
  geom_point(aes(color = is_harvard, size = is_harvard, alpha = is_harvard)) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(sos_sv_rec_data, is_harvard),
    aes(label = player_name),
    color = "#011A50", size = 3.0, max.overlaps = 20,
    box.padding = 0.4, segment.color = "grey60"
  ) +
  facet_wrap(~ role, scales = "free") +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#011A50"),
                     guide = "none") +
  scale_size_manual(values = c("FALSE" = 1.8, "TRUE" = 3.0), guide = "none") +
  scale_alpha_manual(values = c("FALSE" = 0.5, "TRUE" = 1.0), guide = "none") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Serve and reception: raw vs SoS-adjusted PG",
    x = "Raw PG per opportunity",
    y = "SoS-adjusted PG per opportunity"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title.position = "plot",
        strip.text = element_text(face = "bold"))
```

### Interpretation — Serve/Reception SoS

The simplified two-level LMM identifies all four random effects cleanly: the
server effect ($\hat\sigma_\pi = 0.023$) is largest, followed by the two team
effects ($\hat\sigma_{\tilde\tau} = 0.014$, $\hat\sigma_\tau = 0.010$), then
the receiver effect ($\hat\sigma_\rho = 0.009$). The residual SD
($\hat\sigma_\epsilon = 0.160$) dominates, indicating most variation in serve
outcomes is unexplained by player or team identity — which is the same finding
Powers et al. report at NCAA-wide scale.

The qualitative ordering (server $>$ team $>$ receiver) matches the paper.
Single-conference data does not change which actor in the serve/reception
matchup carries the most identifiable signal; it just shrinks the absolute
magnitude of the variance components.

The SoS adjustment shifts Harvard servers slightly downward (Anna Uhr 2.78%
$\to$ 2.49%) and Harvard receivers slightly upward (Anna Uhr 1.25% $\to$
1.59%), consistent with Harvard having faced slightly weaker-than-average
receivers and slightly stronger-than-average servers across the 21-match
schedule. Most shifts are in the 0.001-0.004 range — small in absolute terms,
but methodologically meaningful: the framework runs cleanly at this sample
size and produces interpretable per-player adjustments.

# 10. Strength of Schedule Adjustment — Attack Tree

We extend the SoS framework to the seven attack-tree submodels of Powers et
al. §3.2.2. Each submodel takes the same simplified form (no conference layer):

$$Y_i \sim \mathrm{Normal}\!\left(\alpha^{*} + \tau_{t} + \theta_{a} + \psi_{s}
                                   + \tilde{\tau}_{\tilde t} + \beta_{b}
                                   + \delta_{d},\;
                                   \sigma^2_\epsilon\right)$$

with the digger random effect $\delta_d$ included only on the two outcome
models that involve a digger (`pg_no_block`, `pg_block_through`). The seven
fitted models do double duty: they yield (a) per-contact opponent-side
contributions for SoS adjustment of attack/set/block/dig PG and (b) variance
components used in §11 for individual attribution.

```{r attack-tree-sos-prep}
sos_threshold_att <- 30

attack_data_with_pg <- attack_data_with_pg |>
  dplyr::group_by(player_id) |>
  dplyr::mutate(count = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(player_id_model = ifelse(count < sos_threshold_att,
                                         "0", player_id)) |>
  dplyr::group_by(setter_id) |>
  dplyr::mutate(count = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(setter_id_model = ifelse(count < 100, "0", setter_id)) |>
  dplyr::group_by(blocker_id) |>
  dplyr::mutate(count = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(blocker_id_model = ifelse(count < sos_threshold_att,
                                          "0", blocker_id)) |>
  dplyr::group_by(digger_id) |>
  dplyr::mutate(count = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::mutate(digger_id_model = ifelse(count < sos_threshold_att,
                                         "0", digger_id))
```

```{r attack-tree-sos-fit}
fit_attack_models <- function(data) {
  base_re <- "(1 | team_id_offense) + (1 | team_id_defense) + (1 | player_id_model) + (1 | setter_id_model) + (1 | blocker_id_model)"
  base_re_with_dig <- paste(base_re, "+ (1 | digger_id_model)")

  # is_error model: defense doesn't influence attack errors → drop defense REs
  err_re <- "(1 | team_id_offense) + (1 | player_id_model) + (1 | setter_id_model)"

  models <- list()
  models$is_error <- lme4::lmer(
    as.formula(paste("pg_is_error ~", err_re)),
    data = data,
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  models$is_block <- lme4::lmer(
    as.formula(paste("pg_is_block ~", base_re)),
    data = data |> dplyr::filter(is_no_error),
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  models$is_block_error <- lme4::lmer(
    as.formula(paste("pg_is_block_error ~", base_re)),
    data = data |> dplyr::filter(is_block),
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  models$is_block_return <- lme4::lmer(
    as.formula(paste("pg_is_block_return ~", base_re)),
    data = data |> dplyr::filter(is_block_no_error),
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  models$no_block <- lme4::lmer(
    as.formula(paste("pg_no_block ~", base_re_with_dig)),
    data = data |> dplyr::filter(is_no_block),
    control = lme4::lmerControl(calc.derivs = FALSE, optimizer = "Nelder_Mead")
  )
  models$block_through <- lme4::lmer(
    as.formula(paste("pg_block_through ~", base_re_with_dig)),
    data = data |> dplyr::filter(is_block_through),
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  models$block_return <- lme4::lmer(
    as.formula(paste("pg_block_return ~", base_re)),
    data = data |> dplyr::filter(is_block_return),
    control = lme4::lmerControl(calc.derivs = FALSE)
  )
  return(models)
}

cat("Fitting attack-tree models...\n")
.t0 <- Sys.time()
attack_models <- fit_attack_models(attack_data_with_pg)
cat(sprintf("Done in %.1f seconds.\n", as.numeric(Sys.time() - .t0, units = "secs")))
```

```{r attack-tree-sos-predict}
# For each per-attack PG component, compute the opponent-side contribution.
# Offense-perspective SoS = part of outcome attributable to defense REs.
# Defense-perspective SoS = part of outcome attributable to offense REs.
# Wrap predict() to handle missing REs gracefully (e.g., is_error has no defense REs).
predict_re <- function(model, newdata, re_terms) {
  re_form <- as.formula(paste("~", paste(re_terms, collapse = " + ")))
  available <- names(lme4::ranef(model))
  re_available <- intersect(
    gsub(".*\\| *([^)]+).*", "\\1", re_terms),
    available
  )
  if (length(re_available) == 0) return(rep(0, nrow(newdata)))
  re_form_filtered <- as.formula(paste(
    "~",
    paste(sprintf("(1 | %s)", re_available), collapse = " + ")
  ))
  predict(model, newdata = newdata,
          re.form = re_form_filtered, allow.new.levels = TRUE) |>
    scale(scale = FALSE) |>
    as.numeric()
}

defense_terms <- c("(1 | team_id_defense)", "(1 | blocker_id_model)")
defense_terms_dig <- c("(1 | team_id_defense)", "(1 | blocker_id_model)",
                       "(1 | digger_id_model)")
offense_terms <- c("(1 | team_id_offense)", "(1 | player_id_model)",
                   "(1 | setter_id_model)")

attack_data_with_pg_adj <- attack_data_with_pg |>
  dplyr::mutate(
    sos_is_block_offense        = predict_re(attack_models$is_block,
                                             attack_data_with_pg, defense_terms),
    sos_is_block_defense        = predict_re(attack_models$is_block,
                                             attack_data_with_pg, offense_terms),
    sos_is_block_error_offense  = predict_re(attack_models$is_block_error,
                                             attack_data_with_pg, defense_terms),
    sos_is_block_error_defense  = predict_re(attack_models$is_block_error,
                                             attack_data_with_pg, offense_terms),
    sos_is_block_return_offense = predict_re(attack_models$is_block_return,
                                             attack_data_with_pg, defense_terms),
    sos_is_block_return_defense = predict_re(attack_models$is_block_return,
                                             attack_data_with_pg, offense_terms),
    sos_no_block_offense        = predict_re(attack_models$no_block,
                                             attack_data_with_pg,
                                             defense_terms_dig),
    sos_no_block_defense        = predict_re(attack_models$no_block,
                                             attack_data_with_pg, offense_terms),
    sos_block_through_offense   = predict_re(attack_models$block_through,
                                             attack_data_with_pg,
                                             defense_terms_dig),
    sos_block_through_defense   = predict_re(attack_models$block_through,
                                             attack_data_with_pg, offense_terms),
    sos_block_return_offense    = predict_re(attack_models$block_return,
                                             attack_data_with_pg, defense_terms),
    sos_block_return_defense    = predict_re(attack_models$block_return,
                                             attack_data_with_pg, offense_terms)
  )
```

```{r pg-attack-adj}
# SoS-adjusted aggregations parallel to §8 raw versions.
pg_attack_adj <- attack_data_with_pg_adj |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name, player_position) |>
  dplyr::summarize(
    n = dplyr::n(),
    pg_raw = mean(pg_is_error + pg_is_block + pg_is_block_error +
                    pg_is_block_return + pg_no_block + pg_block_return +
                    pg_block_through),
    pg_adj = mean(
      pg_is_error +
        (pg_is_block        - sos_is_block_offense        * is_no_error) +
        (pg_is_block_error  - sos_is_block_error_offense  * is_block) +
        (pg_is_block_return - sos_is_block_return_offense * is_block_no_error) +
        (pg_no_block        - sos_no_block_offense        * is_no_block) +
        (pg_block_through   - sos_block_through_offense   * is_block_through) +
        (pg_block_return    - sos_block_return_offense    * is_block_return)
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_adj)
```

```{r pg-set-adj}
pg_set_adj <- attack_data_with_pg_adj |>
  dplyr::select(-player_id, -player_name) |>
  dplyr::rename(player_id = setter_id) |>
  dplyr::left_join(dplyr::select(player, player_id, player_name), by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = dplyr::n(),
    pg_raw = mean(pg_is_error + pg_is_block + pg_is_block_error +
                    pg_is_block_return + pg_no_block + pg_block_return +
                    pg_block_through),
    pg_adj = mean(
      pg_is_error +
        (pg_is_block        - sos_is_block_offense        * is_no_error) +
        (pg_is_block_error  - sos_is_block_error_offense  * is_block) +
        (pg_is_block_return - sos_is_block_return_offense * is_block_no_error) +
        (pg_no_block        - sos_no_block_offense        * is_no_block) +
        (pg_block_through   - sos_block_through_offense   * is_block_through) +
        (pg_block_return    - sos_block_return_offense    * is_block_return)
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_adj)
```

```{r pg-block-adj}
# Defense-side: SoS adjustment uses defense-perspective opponent contributions
# (i.e., opponent's offense quality).
pg_block_adj <- attack_data_with_pg_adj |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = blocker_id) |>
  dplyr::left_join(player, by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = sum(is_no_error),
    pg_raw = -weighted.mean(
      pg_is_block + pg_is_block_error + pg_is_block_return +
        pg_no_block / 2 + pg_block_through / 2 + pg_block_return,
      w = is_no_error
    ),
    pg_adj = -weighted.mean(
      (pg_is_block        - sos_is_block_defense        * is_no_error) +
        (pg_is_block_error  - sos_is_block_error_defense  * is_block) +
        (pg_is_block_return - sos_is_block_return_defense * is_block_no_error) +
        (pg_no_block        - sos_no_block_defense        * is_no_block) / 2 +
        (pg_block_through   - sos_block_through_defense   * is_block_through) / 2 +
        (pg_block_return    - sos_block_return_defense    * is_block_return),
      w = is_no_error
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_adj)
```

```{r pg-dig-adj}
pg_dig_adj <- attack_data_with_pg_adj |>
  dplyr::select(-player_id, -player_name, -team_id) |>
  dplyr::rename(player_id = digger_id) |>
  dplyr::left_join(player, by = "player_id") |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n_dig = sum(is_no_block + is_block_through),
    pg_raw = -weighted.mean(
      x = pg_no_block + pg_block_through,
      w = is_no_block + is_block_through
    ),
    pg_adj = -weighted.mean(
      x = (pg_no_block      - sos_no_block_defense      * is_no_block) +
          (pg_block_through - sos_block_through_defense * is_block_through),
      w = is_no_block + is_block_through
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(-pg_adj)
```

```{r adj-overall}
# Roll up to overall PG/set with adjustment applied skill-by-skill.
pg_pass_adj <- pg_reception_adj |>
  dplyr::transmute(player_id, player_name, team_name,
                   n_reception, pg_reception_raw, pg_reception_adj) |>
  dplyr::full_join(
    pg_dig_adj |>
      dplyr::transmute(player_id, player_name, team_name,
                       n_dig, pg_dig_raw = pg_raw, pg_dig_adj = pg_adj),
    by = c("player_id", "player_name", "team_name")
  ) |>
  dplyr::mutate(
    n_reception = dplyr::coalesce(n_reception, 0),
    n_dig = dplyr::coalesce(n_dig, 0),
    n = n_reception + n_dig,
    pg_raw = (n_reception * dplyr::coalesce(pg_reception_raw, 0) +
              n_dig       * dplyr::coalesce(pg_dig_raw, 0)) / n,
    pg_adj = (n_reception * dplyr::coalesce(pg_reception_adj, 0) +
              n_dig       * dplyr::coalesce(pg_dig_adj, 0)) / n
  )

pg_overall_adj <- sets_played |>
  dplyr::full_join(
    pg_serve_adj |>
      dplyr::mutate(pg_serve_raw = n * pg_raw, pg_serve_adj = n * pg_adj) |>
      dplyr::select(dplyr::all_of(id_columns),
                    n_serve = n, pg_serve_raw, pg_serve_adj),
    by = "player_id"
  ) |>
  dplyr::full_join(
    pg_pass_adj |>
      dplyr::mutate(pg_pass_raw = n * pg_raw, pg_pass_adj = n * pg_adj) |>
      dplyr::select(dplyr::all_of(id_columns),
                    n_pass = n, pg_pass_raw, pg_pass_adj),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_set_adj |>
      dplyr::mutate(pg_set_raw = n * pg_raw, pg_set_adj = n * pg_adj) |>
      dplyr::select(dplyr::all_of(id_columns),
                    n_set = n, pg_set_raw, pg_set_adj),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_attack_adj |>
      dplyr::mutate(pg_attack_raw = n * pg_raw, pg_attack_adj = n * pg_adj) |>
      dplyr::group_by_at(id_columns) |>
      dplyr::summarize(n_attack = sum(n),
                       pg_attack_raw = sum(pg_attack_raw),
                       pg_attack_adj = sum(pg_attack_adj),
                       .groups = "drop"),
    by = id_columns
  ) |>
  dplyr::full_join(
    pg_block_adj |>
      dplyr::mutate(pg_block_raw = n * pg_raw, pg_block_adj = n * pg_adj) |>
      dplyr::select(dplyr::all_of(id_columns),
                    n_block = n, pg_block_raw, pg_block_adj),
    by = id_columns
  ) |>
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("^sets_played|^n_|^pg_")),
    function(x) tidyr::replace_na(x, 0)
  )

pg_overall_adj_per_set <- pg_overall_adj |>
  dplyr::mutate_at(
    dplyr::vars(dplyr::matches("^n_|^pg_")),
    function(x, denominator) x / denominator,
    denominator = pg_overall_adj$sets_played
  ) |>
  dplyr::filter(sets_played > 0) |>
  dplyr::mutate(
    n = n_serve + n_pass + n_set + n_attack + n_block,
    pg_raw = pg_serve_raw + pg_pass_raw + pg_set_raw +
             pg_attack_raw + pg_block_raw,
    pg_adj = pg_serve_adj + pg_pass_adj + pg_set_adj +
             pg_attack_adj + pg_block_adj
  ) |>
  dplyr::select(dplyr::all_of(id_columns), sets_played, n, pg_raw, pg_adj,
                dplyr::everything()) |>
  dplyr::arrange(-pg_adj)
```

```{r top10-table}
# Paper Table 5 analog: top 10 by adjusted PG/set with skill-by-skill breakdown
top10_table <- pg_overall_adj_per_set |>
  dplyr::filter(sets_played >= 15, !is.na(player_name)) |>
  dplyr::arrange(dplyr::desc(pg_adj)) |>
  dplyr::slice(1:10) |>
  dplyr::transmute(
    Player = player_name,
    Team = team_name,
    Sets = sets_played,
    `PG/S raw` = pg_raw,
    `PG/S adj` = pg_adj,
    SRV = pg_serve_adj,
    PASS = pg_pass_adj,
    SET = pg_set_adj,
    ATT = pg_attack_adj,
    BLK = pg_block_adj
  )

knitr::kable(
  top10_table,
  caption = "Top 10 players by SoS-adjusted PG per set, with skill-by-skill breakdown (replicates Powers et al. Table 5)",
  digits = 3
)
```

```{r adj-rank-shift-summary}
# Headline: how much do rankings move when SoS is applied?
rank_shift_summary <- pg_overall_adj_per_set |>
  dplyr::filter(sets_played >= 15, !is.na(player_name)) |>
  dplyr::mutate(
    rank_raw = rank(-pg_raw, ties.method = "min"),
    rank_adj = rank(-pg_adj, ties.method = "min"),
    rank_shift = rank_adj - rank_raw
  )

rank_shift_summary |>
  dplyr::summarize(
    n_players           = dplyr::n(),
    cor_pg              = cor(pg_raw, pg_adj),
    mean_abs_rank_shift = mean(abs(rank_shift)),
    median_abs_rank_shift = median(abs(rank_shift)),
    max_abs_rank_shift  = max(abs(rank_shift))
  ) |>
  knitr::kable(caption = "Overall PG/set — rank shift between raw and SoS-adjusted",
               digits = 3)
```

```{r overall-raw-vs-adj-scatter, fig.height = 4}
overall_scatter_data <- pg_overall_adj_per_set |>
  dplyr::filter(sets_played >= 15, !is.na(player_name)) |>
  dplyr::mutate(is_harvard = team_name == "Harvard University")

ggplot(overall_scatter_data, aes(x = pg_raw, y = pg_adj)) +
  geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey85", linewidth = 0.3) +
  geom_point(aes(color = is_harvard, size = is_harvard, alpha = is_harvard)) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(overall_scatter_data, is_harvard),
    aes(label = player_name),
    color = "#011A50", size = 3.0, max.overlaps = 20,
    box.padding = 0.4, segment.color = "grey60"
  ) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#011A50"),
                     guide = "none") +
  scale_size_manual(values = c("FALSE" = 1.8, "TRUE" = 3.0), guide = "none") +
  scale_alpha_manual(values = c("FALSE" = 0.5, "TRUE" = 1.0), guide = "none") +
  scale_x_continuous(labels = number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(
    title = "Overall PG/set: raw vs SoS-adjusted",
    subtitle = sprintf("cor = %.3f; rankings essentially invariant to adjustment within Harvard",
                       cor(overall_scatter_data$pg_raw, overall_scatter_data$pg_adj)),
    x = "Raw PG per set", y = "SoS-adjusted PG per set"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title.position = "plot")
```

### Interpretation — Attack-Tree SoS

The attack-tree SoS adjustment produces a near-identity transformation on
overall PG/set: the correlation between raw and adjusted is 0.997, and within
the 11 Harvard players with $\geq$ 15 sets played, no player's rank changes.

This is not a failure of the framework — it is a finding about the regime in
which the framework is informative. The paper applies SoS to NCAA-wide data
where teams play across conferences with widely varying strengths; opponent
variation is large relative to within-player variation, and SoS produces
substantial corrections (Powers et al. Figure 8 shows visible rank shifts for
many players). On a single 21-match Ivy League schedule, opponent variation
is small, and the seven attack-tree submodels have insufficient cross-team
contrast to distinguish "this attacker faced tough blockers" from "this
attacker is good." The adjustment exists; it is just numerically negligible.

The methodological observation: **SoS adjustment is informative when opponent
variation is large relative to player variation, and degenerate when it is
small.** Future analyses on conference-restricted data should expect
correlations between raw and adjusted PG above 0.99 and rank shifts near
zero.

## Figure 10b. PG/set distribution across all teams (Harvard players highlighted)

A second view of the SoS adjustment, parallel to Powers et al. Figure 8.
Each observation is one player's season-average PG per set across all skills
combined. Reference lines mark Harvard's top performers in the raw and
adjusted distributions.

```{r pg-overall-distribution, fig.height = 4}
overall_dist_data <- pg_overall_adj_per_set |>
  dplyr::filter(sets_played >= 5, !is.na(player_name))

# Harvard reference players
harvard_refs <- overall_dist_data |>
  dplyr::filter(team_name == "Harvard University") |>
  dplyr::arrange(dplyr::desc(pg_adj))

# Top Harvard player for each panel
top_harvard_raw <- harvard_refs |> dplyr::arrange(dplyr::desc(pg_raw)) |>
  dplyr::slice(1)
top_harvard_adj <- harvard_refs |> dplyr::arrange(dplyr::desc(pg_adj)) |>
  dplyr::slice(1)

# Build long-form data for raw vs adjusted faceting
overall_long <- dplyr::bind_rows(
  overall_dist_data |> dplyr::transmute(player_name, team_name,
                                        pg = pg_raw, type = "Raw"),
  overall_dist_data |> dplyr::transmute(player_name, team_name,
                                        pg = pg_adj, type = "SoS-adjusted")
) |>
  dplyr::mutate(type = factor(type, levels = c("Raw", "SoS-adjusted")))

ref_lines <- dplyr::bind_rows(
  top_harvard_raw |> dplyr::transmute(player_name, pg = pg_raw, type = "Raw"),
  top_harvard_adj |> dplyr::transmute(player_name, pg = pg_adj, type = "SoS-adjusted")
) |>
  dplyr::mutate(type = factor(type, levels = c("Raw", "SoS-adjusted")))

ggplot(overall_long, aes(x = pg)) +
  geom_histogram(bins = 30, fill = "#3b5bdb", color = "white",
                 linewidth = 0.3, alpha = 0.7) +
  geom_vline(data = ref_lines,
             aes(xintercept = pg), color = "#011A50",
             linewidth = 0.7, linetype = "solid") +
  geom_text(data = ref_lines,
            aes(x = pg, y = Inf, label = paste(player_name, "(Harvard)")),
            color = "#011A50", angle = 90, hjust = 1.1, vjust = -0.4,
            size = 3.2) +
  facet_wrap(~ type, nrow = 1) +
  labs(
    title = "PG/set distribution: raw vs SoS-adjusted",
    subtitle = "All players in dataset (min 15 sets); Harvard's top performer marked in each panel",
    x = "Points Gained per set",
    y = "Number of players"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", size = 11),
        plot.title.position = "plot")
```

# 11. Variance Decomposition and Individual Attribution

The same seven attack-tree models from §10 carry the variance components for
the paper's attribution method (eqs. 12–14). Below we extract the random-effect
SDs and compute the implied teammate credit shares.

```{r variance-decomposition}
extract_variance_split <- function(models) {
  get_sd <- function(model, term) {
    vc <- lme4::VarCorr(model)
    if (term %in% names(vc)) as.numeric(attr(vc[[term]], "stddev")) else 0
  }
  get_resid <- function(model) attr(lme4::VarCorr(model), "sc")
  get_n     <- function(model) nrow(model@frame)

  tibble::tibble(
    model        = names(models),
    n            = sapply(models, get_n),
    sd_attacker  = sapply(models, get_sd, term = "player_id_model"),
    sd_setter    = sapply(models, get_sd, term = "setter_id_model"),
    sd_blocker   = sapply(models, get_sd, term = "blocker_id_model"),
    sd_digger    = sapply(models, get_sd, term = "digger_id_model"),
    sd_resid     = sapply(models, get_resid)
  ) |>
    dplyr::mutate(
      total_as     = sd_attacker + sd_setter,
      pct_attacker = ifelse(total_as > 0, sd_attacker / total_as, NA),
      pct_setter   = ifelse(total_as > 0, sd_setter   / total_as, NA),
      total_bd     = sd_blocker + sd_digger,
      pct_blocker  = ifelse(total_bd > 0, sd_blocker / total_bd, NA),
      pct_digger   = ifelse(total_bd > 0, sd_digger  / total_bd, NA)
    )
}

variance_split <- extract_variance_split(attack_models)

variance_split |>
  dplyr::select(model, n, sd_attacker, sd_setter,
                sd_blocker, sd_digger, sd_resid) |>
  knitr::kable(
    caption = "Random-effect SDs by attack-tree submodel (paper Appendix Table 9 style)",
    digits = 3
  )

variance_split |>
  dplyr::select(model, pct_attacker, pct_setter, pct_blocker, pct_digger) |>
  knitr::kable(
    caption = "Credit attribution shares (paper Table 4 style)",
    digits = 3
  )
```

```{r variance-components-plot, fig.height = 4.5}
# Make the singular fits visible at a glance: bars per submodel, one per role
vc_long <- variance_split |>
  dplyr::select(model, sd_attacker, sd_setter, sd_blocker, sd_digger,
                sd_resid) |>
  tidyr::pivot_longer(cols = -model, names_to = "role", values_to = "sd") |>
  dplyr::mutate(
    role = factor(
      dplyr::recode(role,
                    sd_attacker = "Attacker",
                    sd_setter   = "Setter",
                    sd_blocker  = "Blocker",
                    sd_digger   = "Digger",
                    sd_resid    = "Residual"),
      levels = c("Attacker", "Setter", "Blocker", "Digger", "Residual")
    ),
    model = factor(model, levels = c(
      "is_error", "is_block", "is_block_error", "is_block_return",
      "no_block", "block_through", "block_return"
    ))
  )

ggplot(vc_long, aes(x = role, y = sd, fill = role)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = ifelse(sd < 0.001, "0", sprintf("%.3f", sd))),
            vjust = -0.3, size = 2.8) +
  facet_wrap(~ model, nrow = 2) +
  scale_fill_manual(values = c(
    "Attacker" = "#3b5bdb", "Setter" = "#9775fa",
    "Blocker" = "#fd7e14", "Digger" = "#20c997",
    "Residual" = "grey60"
  ), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.20))) +
  labs(
    title = "Random-effect SDs across the seven attack-tree submodels",
    subtitle = "Zeros indicate variance components estimated at the singularity boundary (small-sample identifiability)",
    x = NULL, y = "Random-effect SD"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 10),
        panel.grid.major.x = element_blank(),
        plot.title.position = "plot",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

### Interpretation — Variance Decomposition

The seven submodels show systematic small-sample identifiability problems.
Six of the seven models produce at least one variance component estimated at
zero, and `block_return` (n = 750) collapses to zero variance for every
random effect. The pattern is consistent with sample size: submodels with
n $\geq$ 2,000 (`is_block`, `is_error`) identify most of their random
effects, while submodels with n $<$ 1,500 frequently do not.

The most consequential singular fit is the `no_block` model
(n = 2,538), where the attacker variance collapses to zero. Read literally,
this attribution share assigns 0% credit to attackers on clean (non-blocked)
attacks — the most common and most consequential attack outcome. This is
clearly an artifact: the variance decomposition framework cannot identify
attacker-versus-setter contribution at this sample size when both team layers
and a digger random effect are also competing for variance.

We therefore retain shared credit attribution as our primary metric (each
component's PG attributed in full to both attacker and setter) and present
the variance-component attribution as a sensitivity check below. The
framework's attribution machinery requires substantially more data than
25K contacts to estimate reliably.

```{r attribution-comparison}
# Apply the variance-component split to compute "marginal" attacker PG, then
# compare against the shared-credit version (current pipeline default).
get_pct <- function(model_name, role) {
  col <- paste0("pct_", role)
  v <- variance_split[[col]][variance_split$model == model_name]
  ifelse(is.na(v), 0, v)
}

R_att <- list(
  is_error        = get_pct("is_error",        "attacker"),
  is_block        = get_pct("is_block",        "attacker"),
  is_block_error  = get_pct("is_block_error",  "attacker"),
  is_block_return = get_pct("is_block_return", "attacker"),
  no_block        = get_pct("no_block",        "attacker"),
  block_through   = get_pct("block_through",   "attacker"),
  block_return    = get_pct("block_return",    "attacker")
)

attack_attributed <- attack_data_with_pg |>
  dplyr::mutate(
    pg_attacker_marg =
      pg_is_error        * R_att$is_error +
      pg_is_block        * R_att$is_block +
      pg_is_block_error  * R_att$is_block_error +
      pg_is_block_return * R_att$is_block_return +
      pg_no_block        * R_att$no_block +
      pg_block_through   * R_att$block_through +
      pg_block_return    * R_att$block_return
  )

pg_attacker_compare <- attack_attributed |>
  dplyr::left_join(team, by = "team_id") |>
  dplyr::group_by(player_id, player_name, team_name) |>
  dplyr::summarize(
    n = dplyr::n(),
    pg_shared = sum(pg_is_error + pg_is_block + pg_is_block_error +
                    pg_is_block_return + pg_no_block + pg_block_through +
                    pg_block_return),
    pg_marg = sum(pg_attacker_marg),
    .groups = "drop"
  ) |>
  dplyr::filter(n >= 30, !is.na(player_name)) |>
  dplyr::mutate(
    rank_shared = rank(-pg_shared, ties.method = "min"),
    rank_marg = rank(-pg_marg, ties.method = "min"),
    rank_shift = rank_marg - rank_shared
  ) |>
  dplyr::arrange(-pg_marg)

pg_attacker_compare |>
  dplyr::summarize(
    n_players = dplyr::n(),
    cor_pg = cor(pg_shared, pg_marg, use = "complete.obs"),
    mean_abs_rank_shift = mean(abs(rank_shift)),
    max_abs_rank_shift = max(abs(rank_shift))
  ) |>
  knitr::kable(
    caption = "Attacker rankings — shared-credit vs marginal attribution",
    digits = 3
  )
```

Despite the singular fits, the attribution comparison is not numerically
trivial: the marginal attribution shifts attacker rankings by 2.8 positions
on average and up to 14 positions across the 38-player sample. The choice of
attribution method is not a robustness check that washes out — it
materially affects who comes out on top of the rankings. This reinforces
the case for treating attribution as a methodological choice that matters,
not a detail that can be papered over.

# 12. Bootstrap Standard Errors (optional)

The paper computes SEs by resampling matches with replacement. This is
expensive (~minutes per resample on full data). The chunk below is scaffolded
but disabled by default — set `eval = TRUE` to run.

```{r bootstrap, eval = FALSE}
set.seed(42)
n_boot <- 200
match_ids <- unique(plays$match_id)
boot_results <- vector("list", n_boot)

for (b in seq_len(n_boot)) {
  if (b %% 20 == 0) cat(sprintf("  Bootstrap %d / %d\n", b, n_boot))
  # Resample matches, refit models, store variance_split.
  # See code_for_paper.R for the full body (~120 lines).
}
```

# 13. Save Outputs

```{r save-outputs}
# Raw PG tables
pg_overall_per_set |> dplyr::filter(sets_played >= 15) |>
  write.csv("pg_overall.csv", row.names = FALSE, na = "")
pg_serve     |> dplyr::filter(n >= 20)            |> write.csv("pg_serve.csv",     row.names = FALSE, na = "")
pg_pass      |> dplyr::filter(n >= 30)            |> write.csv("pg_pass.csv",      row.names = FALSE, na = "")
pg_set       |> dplyr::filter(n >= 100)           |> write.csv("pg_set.csv",       row.names = FALSE, na = "")
pg_attack    |> dplyr::filter(n >= 30)            |> write.csv("pg_attack.csv",    row.names = FALSE, na = "")
pg_block     |> dplyr::filter(n >= 30)            |> write.csv("pg_block.csv",     row.names = FALSE, na = "")

# SoS-adjusted PG tables
pg_overall_adj_per_set |> dplyr::filter(sets_played >= 15) |>
  write.csv("pg_overall_adj.csv", row.names = FALSE, na = "")
pg_serve_adj     |> dplyr::filter(n >= 20)            |> write.csv("pg_serve_adj.csv",     row.names = FALSE, na = "")
pg_reception_adj |> dplyr::filter(n_reception >= 30)  |> write.csv("pg_reception_adj.csv", row.names = FALSE, na = "")
pg_attack_adj    |> dplyr::filter(n >= 30)            |> write.csv("pg_attack_adj.csv",    row.names = FALSE, na = "")
pg_set_adj       |> dplyr::filter(n >= 100)           |> write.csv("pg_set_adj.csv",       row.names = FALSE, na = "")
pg_block_adj     |> dplyr::filter(n >= 30)            |> write.csv("pg_block_adj.csv",     row.names = FALSE, na = "")
pg_dig_adj       |> dplyr::filter(n_dig >= 30)        |> write.csv("pg_dig_adj.csv",       row.names = FALSE, na = "")

# Variance and attribution
serve_var_components |> write.csv("serve_var_components.csv", row.names = FALSE, na = "")
variance_split       |> write.csv("variance_split.csv",       row.names = FALSE, na = "")
pg_attacker_compare  |> write.csv("pg_attacker_compare.csv",  row.names = FALSE, na = "")

cat("All outputs saved.\n")
```
